---
title: "Swift Parrot Data Cleaning"
subtitle: "Atlas of Living Australia — Occurrence Records"
author: "Farzan Momayezi"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

This script cleans the raw Swift Parrot (*Lathamus discolor*) occurrence records
downloaded from the [Atlas of Living Australia](https://www.ala.org.au/).
Irregularities were first detected visually in Tableau (spatial outliers,
inconsistent date formats and extreme coordinate uncertainty) and are corrected
here programmatically so the pipeline is fully reproducible.

Paths are **relative to the project root**, so the script runs on any machine
without editing. Open `R/Data_wrangling.Rproj` in RStudio (or set the working
directory to the repo root) before running.

## Dependencies

```{r libraries}
library(dplyr)      # data manipulation
library(lubridate)  # robust date parsing
```

## 1. Load the raw dataset

```{r load}
data <- read.csv(
  file.path("..", "data", "raw", "ALA_S12026PE1.csv"),
  stringsAsFactors = FALSE
)

cat("Raw records:", nrow(data), "\n")
```

## 2. Fix date formats and drop future / invalid dates

The `observationDate` field mixes several formats (`d/m/Y`, `Y-m-d`, `m/d/Y`).
`lubridate::parse_date_time()` standardises them; records that fail to parse or
fall in the future are removed to keep the temporal analysis sound.

```{r dates}
data$observationDate <- parse_date_time(
  data$observationDate,
  orders = c("d/m/Y", "Y-m-d", "m/d/Y")
)

cleaned_data <- data %>%
  filter(!is.na(observationDate) & observationDate <= Sys.Date())
```

## 3. Correct geographic outliers (points in the ocean)

Some records had latitude and longitude recorded in the wrong fields, placing
Swift Parrots in the Indian Ocean / Western Australia. The species is endemic to
south-eastern Australia and Tasmania, so positive-latitude points are swapped
back into the correct axes.

```{r location}
cleaned_data <- cleaned_data %>%
  mutate(
    temp_lat        = decimalLatitude,
    decimalLatitude = ifelse(decimalLatitude > 0, decimalLongitude, decimalLatitude),
    decimalLongitude = ifelse(temp_lat > 0, temp_lat, decimalLongitude)
  ) %>%
  select(-temp_lat)
```

## 4. Correct state labels for Tasmania

Points that fall inside Tasmania's geographic bounds but were mislabelled
(e.g. as New South Wales) are relabelled.

```{r state}
cleaned_data <- cleaned_data %>%
  mutate(
    stateProvince = ifelse(
      decimalLatitude < -40 & decimalLongitude > 143,
      "Tasmania",
      stateProvince
    )
  )
```

## 5. Remove coordinate-uncertainty outliers

Records with a positional uncertainty above 10 km are too vague for habitat
mapping and are dropped.

```{r uncertainty}
cleaned_data <- cleaned_data %>%
  filter(coordinateUncertaintyInMeters <= 10000)
```

## 6. Save the cleaned dataset

```{r save}
out_path <- file.path("..", "data", "processed", "Final_Cleaned_Swift_Parrot_Data.csv")
write.csv(cleaned_data, out_path, row.names = FALSE)

cat("Cleaned records:", nrow(cleaned_data), "\n")
cat("Saved to:", out_path, "\n")
```
